---
title: 'EEMB 140 -- Lab 3: Leaf Traits'
author: 'your name here'
output:
  html_document: default
date: "January XX, 2022"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Install and load packages
```{r}
# install.packages("tidyverse")

library(tidyverse)
library(ggplot2)

```

# 2. Import your data in a csv file
After all groups have finished entering trait data, download the group data sheet on Google Drive as a csv, and use the read.csv() function to load the data

```{r}

class_data <- read.csv("insert name of file here")

```

# 3. Graphs & Interpretations

## a. Leaf Area

### Leaf area by species
```{r}
ggplot(class_data, aes(x = Species, y = avg_leaf_area_cm2)) +
  geom_boxplot(aes(fill = Species)) + # adding color
  ggtitle("Leaf Area (cm2) by Species") # adding a title

```

Questions 1: Using the graph above, were there any differences between the species, and why do you think there is? What might these values of leaf area tell us about the plants' ecology and physiology?
(write your answer here)

### Leaf area by direction?
```{r}

ggplot(class_data, aes(x = Direction, y = avg_leaf_area_cm2)) +
  geom_boxplot(aes(fill = Direction)) + # adding color
  ggtitle("Leaf Area (cm2) by Direction") # adding a title

```

Question 2: Did we see any effect of the direction (North/South/East/West) across all species? Why do you think that is?
(write your answer here)

## b. LMA(Leaf Mass per area g/cm2)

```{r}


```
Question 3: What trends did you observe in LMA, and what does this mean ecologically?
(write your answer here)

## c. LWC (Leaf water content, g/g)

```{r}

```
Question 4: What trends did you observe LWC, and what does this mean ecologically?
(write your answer here)


## d. Al:As (Leaf Area to Stem Area, cm2/mm2)
```{r}


```
Question 5: What trends did you observe LWC, and what does this mean ecologically?
(write your answer here)

## Homework

1. If you were to re-design the sampling methods based on your results, would you make any changes? Why or why not?
(write 2-3 sentences here)

2. Choose one of the species we covered in lab, and look up information on the habitats that they live in. Do the results you observed in the trait data match your expectations for their ecological range? Why or why not?
(write 2-3 sentences here)


