---
title: 'EEMB 140L -- Lab 5: Rosemary Growth Data Analysis'
author: 'your name here'
output:
  html_document: default
date: "January 26, 2023"
---

# Note: you will not be turning in a knitted RMarkdown file for this lab, just the printed lab writeup. This code is just to provide a jumping-off point.

* Import packages (this code chunk will not knit properly but it will import some packages for you)
```{r, include = FALSE}
install.packages("ggplot2")
install.packages("dplyr")
library(ggplot2)
library(dplyr)
```     
* Set the working directory (you can also do this by clicking "Session" above) and import CSV file with your data:
```{r}
setwd("FOLDER_ADDRESS_HERE")
data <- read.csv("YOUR_FILE_HERE.csv")
```

* Example plot: point/linear regression plot of the amount of flowers versus the shadiness of each bush (you'll need to make sure the names in this chunk match the names in your data file!)
```{r}
ggplot(data, aes(x = shade, y = flowers, color = species))+
  geom_point()+
  geom_smooth(method = "lm")
```