---
title: "Lab_10_Carbon_stocks"
output: html_document
date: "2023-02-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# helpful packages
library(tidyverse)
library(ggplot2)

```

## Carbon Stocks

In this Markdown, you will be working with your tree stand data to...


1) Make sure your group's tree stand survey data is completely entered

2) Read in your data here:
```{r}

# stand_data <- read.csv("title.csv")

stand_data <- read.csv("stand_data_test.csv")

```

3) Calculate tree stand density for each plot

First, you need to calculate the area that you surveyed, which was a circle with a 5 meter diameter:
```{r}

survey_area <- pi * 5^2

```

Then, you can get a count of the total number of trees in each plot using...
```{r}

tree_density <- stand_data %>% 
  group_by(survey_number) %>%  # first, group by each survey
  count() %>% # then get total number of trees in each survey
  mutate(density = n / survey_area)# this adds a new column called "density" that divides the count column by our survey area

```

Qustions 1) Look at this new dataset called "tree_density", do all the plots seem similar in density or do any stand out?
(write answer here)


