---
title: "Lab_10_Carbon_stocks"
output: html_document
date: "2023-02-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# helpful package:
library(tidyverse)

```

## Carbon Stocks

In this Markdown, you will be working with uour tree stand data to estimate different forestry metrics, as well as work with allometric equations and tree carbon data to estimate tree stand carbon.

### 1) Make sure your group's tree stand survey data is completely entered

### 2) Read in your data here:
```{r}

# stand_data <- read.csv("title.csv")

```

### 3) Calculate tree stand density for each plot

To calculate tree stand density, you will first need to calculate the area that was surveyed, which was a circle with a 5 meter diameter:
```{r}

# A = pi*r^2


```

Then, you can get a count of the total number of trees in each plot using. First, group by survey (survey_name column), and use the count() function to get a total number of trees in each.  You can then use the mutate() function to caculate the density, and mulitply by 10,000 to convert it to trees/hectare.
```{r}

tree_density <- stand_data %>% 
  %>%  # first, group by each survey
  %>% # then get total number of trees in each survey
  # this adds a new column called "density" that divides the count column by our survey area and multiples by 10,000 to get trees per hectare

```

### Question 1) Look at this new data set called "tree_density", do all the plots seem similar in density or do any stand out?
(write answer here)

### 4) Calculate the stand basal area (cm2 stem area per hectare)

Another way to think about trees in a stand rather than just their count, is the area that is occupied by their trunk.

There are lots of ways to do this! One way is to first create a new column calculating each trees basal area, that is, use the diameter to calculate the area. Then group again by survey_number. Next, create another column called stand_basal_area where you divide the mean basal_area column by the survey area and multiply by 10,000 to get the basal area per hectare

```{r}
tree_basal_area <- stand_data %>% 
   %>% # create new column
   %>% # group by plot
   # create new column for mean basal area, convert into total basal area per hectare

```

### Question 2: Do you think stand basal area might be more accurate than tree number when thinking about ecological impacts? Why or why not?
(answer here)


### 5) Calculate average stand height

Another piece of the puzzle to get accurate biomass estimates is using tree height.  In order to get average stand tree heights, we first need to create a new column where using the mutate function. Remember to calculate height, this is done by multiplying the degrees (make sure to divide by 100) by the distance, but don't forget to add the height of your eyes! Then you can again group by survey and use the summarise() function to calculate the mean
```{r}
stand_height <- stand_data %>% 
   # create new column calculating height
   %>% # group by plot
   # find mean tree height

```

### Challenge graph: make a boxplot of tree height by plot

To do this, you will want to use the tree_id column

```{r}



```


### 6) Use an allometric equation to plot standing biomass

In order to determine the above ground dry biomass of our eucs, we could simply treat them as a column using our calculations for height and diameter; however, trees are a little bit more complicated than that.  Instead, researches have determined more accurate and sometimes species-specific allometric equations. For eucs, we will be using the AGB = .00213 * DBH_cm^1.829, which other researchers determined for a similar species of eucalyptus.

So we can simply make a new column for above-ground biomass (in kg) using the mutate() function, and then add this up for each survey using the summarise() function.
```{r}
tree_biomass_eucs <- stand_data %>%
  %>% # create new column to calculate abg
   %>% # group by survey
   # get total biomass per plot and convert to per hectare

```

### 7) Calculate above-ground carbon

Using our estimates for above-ground dry biomass, we can now determine the amount of carbon stored within each survey site

To do this, we will multiply the dry biomass by 35%, which is approximately the amount of carbon present that researchers found have found when processing eucalyptus tree core samples.

```{r}
tree_carbon <- tree_biomass_eucs %>% 
   # create new colum for total biomass by multiplying by .35

```

### 8) Oak tree above-ground carbon storage
We will now do the same work as above for a similar stand of trees, but imagine that they are oaks instead of eucalyptus. How do you think this might affect carbon storage?

To do this, use your DBH calculations but with the allometric equation ABG = 0.0000062 * DBH_cm^2.584 and the percent carbon as 45%

```{r}

tree_biomass_oaks <- stand_data %>%
  %>% # new column for euc abg
  %>% # new column for oak abg
  %>%  # group by survey plot
  %>% # get total abg for each, and convert to per hectare value
  %>% # convert biomass to carbon
  # convert biomass to carbon

```

### Question 3:Which species stores more abovegound carbon? Given that eucalypus are invasive species, what do you think this means for balancing their management? What are some of their costs and benefits to an ecosystem?

(write at least 3 sentences here)

### Challenge Question: Create a boxplot/histogram/violin plot comparing oak and eucalyptus aboveground carbon


